---
title: "When do the big waves come? A year on two neighbouring wave buoys"
subtitle: "Assignment 4 ETC5512"
author: "Hanwen Liu"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 4.6
    fig-align: center
    fig-cap-location: bottom
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

```{r}
#| label: setup
#| include: false

source("R/02_clean_process.R", local = knitr::knit_global())

library(tidyverse)
library(scales)
library(sf)
library(ozmaps)

theme_set(
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey"),
          strip.text = element_text(face = "bold"))
)
season_cols <- c(Summer = "red", Autumn = "yellow",
                 Winter = "blue", Spring = "green")
site_cols   <- c("Gold Coast" = "lightblue", "Tweed Heads" = "pink")

buoys <- tibble(
  site = factor(c("Gold Coast", "Tweed Heads"), levels = levels(waves$site)),
  lon  = c(153.44095, 153.576333),
  lat  = c(-27.9649,  -28.177083),
  depth_m = c(17, 22)
)
haversine <- function(lo1, la1, lo2, la2) {
  R <- 6371; dlat <- (la2 - la1) * pi/180; dlon <- (lo2 - lo1) * pi/180
  a <- sin(dlat/2)^2 + cos(la1*pi/180) * cos(la2*pi/180) * sin(dlon/2)^2
  2 * R * asin(sqrt(a))
}
dist_km <- haversine(buoys$lon[1], buoys$lat[1], buoys$lon[2], buoys$lat[2])

gc         <- waves   %>% 
  filter(site == "Gold Coast")
monthly_gc <- monthly %>%  
  filter(site == "Gold Coast")

warm  <- monthly_gc %>% 
  slice_max(mean_sst, n = 1)
cool  <- monthly_gc %>% 
  slice_min(mean_sst, n = 1)
rough <- monthly_gc %>% 
  slice_max(mean_hs,  n = 1)
calm  <- monthly_gc %>% 
  slice_min(mean_hs,  n = 1)
sst_swing <- warm$mean_sst - cool$mean_sst

n_big3   <- sum(gc$hs >= 3, na.rm = TRUE)
pct_big3 <- 100 * mean(gc$hs >= 3, na.rm = TRUE)
top_hs   <- gc %>% 
  slice_max(hs,   n = 1)   
top_hmax <- gc %>% 
  slice_max(hmax, n = 1)   
mar_miss <- monthly_gc$pct_missing_hs[monthly_gc$month == "Mar"]

cor_hs_tp  <- cor(gc$hs, gc$tp,  use = "complete.obs")
cor_hs_sst <- cor(gc$hs, gc$sst, use = "complete.obs")
cor_month  <- cor(monthly_gc$mean_hs, monthly_gc$mean_sst)

site_stats <- waves %>% 
  group_by(site) %>% 
  summarise(mean_hs = mean(hs, na.rm = TRUE),
            mean_sst = mean(sst, na.rm = TRUE),
            max_hmax = max(hmax, na.rm = TRUE), .groups = "drop")
gc_mhs  <- site_stats$mean_hs[site_stats$site == "Gold Coast"]
tw_mhs  <- site_stats$mean_hs[site_stats$site == "Tweed Heads"]
gc_hmax <- site_stats$max_hmax[site_stats$site == "Gold Coast"]
tw_hmax <- site_stats$max_hmax[site_stats$site == "Tweed Heads"]

wide <- waves %>% 
  mutate(s = recode(site, "Gold Coast" = "gc", "Tweed Heads" = "tw")) %>% 
  select(s, datetime, hs, sst) %>% 
  pivot_wider(names_from = s, values_from = c(hs, sst))
hsp  <- wide %>% 
  filter(!is.na(hs_gc),  !is.na(hs_tw))
sstp <- wide %>% 
  filter(!is.na(sst_gc), !is.na(sst_tw))
cor_hs_sites  <- cor(hsp$hs_gc,  hsp$hs_tw)
cor_sst_sites <- cor(sstp$sst_gc, sstp$sst_tw)
n_matched     <- nrow(hsp)
mar_tw <- monthly %>% 
  filter(site == "Tweed Heads", month == "Mar")

states <- ozmaps::ozmap_states

fmt1 <- function(x) formatC(x, format = "f", digits = 1)
fmt2 <- function(x) formatC(x, format = "f", digits = 2)
```

::: panel-tabset
<!-- ===================================================================== -->

<!-- TASK 1 -->

## Data and Documenting

### Q1 — The question

This blog post answers, from a full year of half-hourly wave-buoy records:

> **How did wave conditions and sea temperature change across 2025 off the Gold Coast, when did the biggest-wave events arrive, and how does wave height relate to wave period (swell versus local wind-sea) — and does a second buoy 27 km south tell the same story?**

I chose this question because it is tightly scoped — one year, two open datasets — yet rich enough to exercise a wide range of the unit's skills: datetime parsing, sentinel-value cleaning, aggregating a 17,520-row half-hourly series, joining two datasets on a common time grid, and time-series, seasonal, spatial and correlation analysis. Adding the second buoy is deliberate: rather than just describing one record, it lets me *test* whether the patterns are real signals in the ocean or merely one instrument's quirks.

### Q2 — The data {#sec-data}

I use **two** datasets from the **Queensland Government Open Data Portal**, both the "Wave data – 2025" resource of a *Coastal Data System – Waves* dataset, both **observational sensor data** from a **Datawell Directional Waverider** buoy recording a 26.6-minute sample every **30 minutes**.

|   | **Dataset 1 — Gold Coast** (primary) | **Dataset 2 — Tweed Heads** (comparison) |
|------------------------|------------------------|------------------------|
| Source / custodian | DETSI + City of Gold Coast | DETSI + Tweed River Sand Bypassing |
| Location | 27.965° S, 153.441° E, 17 m depth | 28.177° S, 153.576° E, 22 m depth |
| Rows (2025) | 17,520 | 17,520 (after clipping to 2025) |
| Licence | CC BY 4.0 | CC BY 4.0 |
| Portal | [link](https://www.data.qld.gov.au/dataset/coastal-data-system-waves-gold-coast) | [link](https://www.data.qld.gov.au/dataset/coastal-data-system-waves-tweed-heads) |

**Type.** Neither is a survey or an experiment; each is effectively a **census** of every valid half-hourly record the buoy made in 2025 — a complete, regular time grid.

**Why this suits the question.** Each file carries exactly the variables I need — significant and maximum wave height (`Hs`, `Hmax`), two wave periods (`Tz`, `Tp`), peak wave direction, and sea surface temperature (`SST`). One buoy answers the "when" questions; the **second, 27 km south, is what lets me check whether the answers are real**. Both are \~1 MB and openly licensed, so the work is fully reproducible.

**Limitations.**

- **Missing data is coded, not blank.** Failed readings are the sentinel `-99.9`; untreated it would poison every average. Gold Coast is \~4.5% missing for wave parameters — but **not evenly**: **March alone is \~`r fmt1(mar_miss)`% missing** (an outage). Tweed Heads is far more complete (\~0.1%), which turns out to matter.
- **Two buoys, one year.** Findings describe these sites in 2025, not a long-term climatology.
- **Point in space.** A buoy measures open-water sea state, not the breaking surf at any particular beach.

**Ethics & privacy.** Both datasets measure the physical ocean only — **no people, vessels or identifiable subjects** — so privacy/re-identification risk is negligible. The obligations are *attribution* (CC BY 4.0) and *responsible interpretation*: this is a curiosity project, **not** a marine-safety product.

### Q3 — Downloading and processing the data

**Download (`R/01_download.R`).** The script downloads each buoy's CSV from its portal URL and **falls back to the committed copy in `data/raw/`** if the portal is unreachable, so the pipeline reproduces online *or* offline. Raw files are never edited by hand.

```{r}
#| eval: false
#| code-fold: false

ok <- tryCatch({ download.file(s$url, s$dest, mode = "wb"); TRUE },
               error = function(e) FALSE, warning = function(w) FALSE)
if (!ok && !file.exists(s$dest)) stop("Download failed and no committed copy.")
header <- strsplit(readLines(s$dest, n = 1), ",")[[1]]
stopifnot(identical(header, expected_cols))
```

**Process (`R/02_clean_process.R`).** The core steps — applied to both buoys:

```{r}
#| eval: false
#| code-fold: false

read_buoy <- function(path, site) {
  read_csv(path, col_types = cols(`Date/Time (AEST)` = col_character(),
                                  .default = col_double())) %>% 
    mutate(site = site, .before = 1)
}
waves <- bind_rows(read_buoy(gc_path, "Gold Coast"),
                   read_buoy(tw_path, "Tweed Heads")) %>% 
  mutate(datetime = ymd_hms(`Date/Time (AEST)`, truncated = 1, tz = "Australia/Brisbane")) %>% 
  mutate(across(c(hs, hmax, tz, tp, dir, sst), \(x) if_else(x == -99.9, NA, x))) %>% 
  filter(year(datetime) == 2025)  
```

The pipeline then runs **per-site assertions** (each site: 17,520 rows, every timestamp parsed, no duplicates, a regular 30-minute grid) and writes two **processed** datasets: `waves_2025_clean.csv` (tidy half-hourly, both sites, sentinel → `NA`) and `waves_2025_monthly.csv` (site × month summary). Raw and processed data are both saved under `data/`.

### Q4 — Data documentation for markers

Full documentation lives alongside the data:

- **`docs/metadata.md`** — provenance, custodians, licences, coverage, instruments, ethics (both buoys).
- **`docs/data_dictionary.csv`** — every variable with units, valid 2025 range and missing code.
- **`README.md`** — folder map and one-command reproduction steps.

A **sample of a raw file** (note the `-99.9` sentinel and the half-hourly step):

```{r}
#| code-fold: true
#| code-summary: "Show code"
read_csv("data/raw/gold-coast-mk4-2025-waveverifieddata.csv", n_max = 3,
         col_types = cols(.default = col_character())) |>
  knitr::kable(caption = "First 3 rows of the Gold Coast raw CSV.")
```

**Missingness by month and site** — the one big gap is Gold Coast in March; Tweed Heads is near-complete all year (this is what makes it a useful cross-check):

```{r}
#| label: tbl-missing
#| code-fold: true
#| fig-height: 3.0
#| fig-cap: "Percentage of half-hours with no valid Hs reading, by month and buoy. Gold Coast's ~25% March outage is the one large gap; Tweed Heads is near-complete all year."
monthly |>
  ggplot(aes(month, pct_missing_hs, fill = site)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = site_cols, name = NULL) +
  labs(x = NULL, y = "% missing Hs",
       title = "Where are the gaps?  Missing wave readings by month, 2025") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme(legend.position = "top")
```

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### Reading the ocean's diary, one buoy at a time

Every thirty minutes, a yellow buoy bobbing off the Gold Coast writes a single line in a diary: how big the waves are, how far apart, which way they're marching, and how warm the sea is. Do that for a year and you get **17,520 lines** — a complete record of one patch of ocean through 2025. I wanted to read that diary and answer three things I've always wondered about as someone who watches the coast: *when is the sea roughest, when do the genuinely big swells turn up, and what does the data say about "swell" versus messy local "wind-sea"?* Then I do the honest thing and ask a fourth: **is any of it real, or just this one buoy?**

#### Two rhythms that don't keep time together

The first thing the year reveals is that the ocean keeps **two different clocks**.

```{r}
#| label: fig-rhythms
#| fig-cap: "Two seasonal rhythms off the Gold Coast in 2025. Sea temperature (top) follows a smooth solar cycle that lags the calendar — warmest in late summer, coldest in late winter. Wave height (bottom) does not: it peaks in autumn and has a sharp winter spike, and is calmest in spring. Points are coloured by Southern-Hemisphere season."
#| fig-height: 5.4
monthly_gc %>% 
  transmute(month, season,
            `Mean significant wave height (m)` = mean_hs,
            `Mean sea surface temperature (°C)` = mean_sst) %>% 
  pivot_longer(-c(month, season), names_to = "metric", values_to = "value") %>% 
  ggplot(aes(month, value, group = 1)) +
  geom_line(colour = "grey", linewidth = 0.7) +
  geom_point(aes(colour = season), size = 3) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = season_cols, name = "Season") +
  labs(x = NULL, y = NULL,
       title = "The sea's temperature and its roughness follow different calendars") +
  theme(legend.position = "top")
```

**Sea temperature** behaves exactly as you'd hope a thermometer would: a smooth wave from a warm **`r warm$month` (`r fmt1(warm$mean_sst)`°C)** down to a cold **`r cool$month` (`r fmt1(cool$mean_sst)`°C)** — a swing of about **`r fmt1(sst_swing)`°C** across the year. Notice it *lags the calendar*: the warmest month is late summer, not the summer solstice, and the coldest is late winter. That delay is the ocean's huge thermal inertia — water is slow to heat and slow to cool.

**Wave height keeps a stubbornly different beat.** The roughest month is **`r rough$month` (mean Hs `r fmt2(rough$mean_hs)` m)** and the calmest is **`r calm$month` (`r fmt2(calm$mean_hs)` m)**. Seas build through autumn, throw a spike in late winter, then go quiet in spring — *not* tracking temperature at all. Waves are made by **wind and distant storms**, which have their own seasonal calendar, so the two curves drift in and out of step. That mismatch is the whole point of the chart, and it sets up a trap I'll spring at the end.

#### When do the big waves actually come?

A monthly average hides the drama. If I instead take the **biggest wave each day** and mark every half-hour where significant height topped a stormy **3 m** (only **`r n_big3`** half-hours all year, `r fmt2(pct_big3)`% of the time), the big events don't dribble out evenly — they **clump**.

```{r}
#| label: fig-events
#| fig-cap: "Daily maximum significant wave height at the Gold Coast through 2025. Most days sit below 2 m; the genuinely big seas (red, Hs ≥ 3 m) cluster into a few multi-day events, overwhelmingly in autumn (early March) and a sharp burst in late winter (early August)."
gc %>% 
  group_by(date) %>% 
  summarise(max_hs = max(hs, na.rm = TRUE), .groups = "drop") %>% 
  mutate(max_hs = ifelse(is.finite(max_hs), max_hs, NA_real_)) %>% 
  ggplot(aes(date, max_hs)) +
  geom_hline(yintercept = 3, linetype = "dashed", colour = "grey") +
  geom_line(colour = "grey", linewidth = 0.4) +
  geom_point(data = ~ filter(.x, max_hs >= 3), colour = "red", size = 1.8) +
  annotate("text", x = as.Date("2025-03-05"), y = 4.7, label = "early March",
           colour = "red", size = 3.4, fontface = "bold") +
  annotate("text", x = as.Date("2025-08-03"), y = 4.9, label = "early August",
           colour = "red", size = 3.4, fontface = "bold") +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = NULL, y = "Daily max Hs (m)",
       title = "Big waves don't trickle in — they arrive in bunches")
```

Two events dominate the year, and the lovely twist is that they sit in **opposite seasons at opposite sea temperatures**:

- **Early March (warm, autumn).** This event produced the year's **largest single wave**, an `Hmax` of **`r fmt2(top_hmax$hmax)` m** on **`r format(top_hmax$datetime, "%d %B")`**, while the sea was still a balmy **`r fmt1(top_hmax$sst)`°C** — the signature of a warm east-coast low / ex-tropical system. Gold Coast's \~25% March outage means we're probably *under*-counting this one (more on that below).
- **Early August (cold, winter).** This produced the year's **highest significant height**, **Hs `r fmt2(top_hs$hs)` m** on **`r format(top_hs$datetime, "%d %B")`**, with the sea down at **`r fmt1(top_hs$sst)`°C** — a classic cold-season, long-period swell.

So "big-wave season" in 2025 wasn't a single season at all: the two heavyweights came from completely different weather, half a year and \~6.5°C apart.

#### Swell or slop? What the wave period tells us

Surfers care less about raw height than about **period** — long, orderly *swell* from distant storms versus short, choppy *wind-sea* from local weather. Plotting every half-hour of height against peak period separates the two:

```{r}
#| label: fig-swell
#| fig-cap: "Every valid half-hour of Gold Coast significant wave height vs peak period (colour = number of records, log scale). The cloud has two homes: short-period local wind-sea on the left, and a long-period swell band around 9–14 s on the right where the biggest seas live. Beyond ~14 s sit long but low waves — tired swell that has travelled a long way."
gc %>% 
  filter(!is.na(hs), !is.na(tp)) %>% 
  ggplot(aes(tp, hs)) +
  geom_bin2d(bins = 45) +
  scale_fill_viridis_c(trans = "log10", name = "records", labels = label_number()) +
  annotate("rect", xmin = 2, xmax = 8, ymin = 0, ymax = 4.7,
           fill = NA, colour = "grey", linetype = "dotted") +
  annotate("text", x = 5, y = 4.5, label = "wind-sea\n(short period)",
           size = 3.2, colour = "grey", lineheight = 0.9) +
  annotate("text", x = 13.5, y = 4.5, label = "swell\n(long period)",
           size = 3.2, colour = "grey", lineheight = 0.9) +
  labs(x = "Peak period Tp (s)", y = "Significant wave height Hs (m)",
       title = "Two kinds of sea: choppy wind-sea and long-marching swell")
```

The relationship is real but **weak** (correlation `r fmt2(cor_hs_tp)`): longer-period seas tend to be bigger, yet plenty of long-period records are small. Binning confirms it — mean height climbs from \~0.9 m at the shortest periods to \~1.35 m in the **12–14 s** swell band, then *falls back* for the very longest periods (\> 14 s). Those longest waves are **decayed swell from far-off Southern Ocean storms**: long and orderly, but worn down to a low height by the journey. The biggest local seas, by contrast, ride the **9–14 s** band — long enough to pack energy, close enough to still be tall.

#### One ocean, two buoys: is any of this real?

Everything above rests on a single yellow float. A fair worry: what if the "story" is just *that* instrument? So I bring in a neighbour — the **Tweed Heads** buoy, **`r fmt1(dist_km)` km** south, just across the NSW border.

```{r}
#| label: fig-map
#| fig-cap: "The two wave buoys: Gold Coast (QLD) and Tweed Heads (NSW), about 27 km apart on the same stretch of coast. State outlines from the ozmaps package."
#| fig-height: 4.8
#| fig-width: 6.6
ggplot(states) +
  geom_sf(fill = "white", colour = "grey", linewidth = 0.3) +
  annotate("text", x = 152.95, y = -27.7,  label = "QUEENSLAND",
           colour = "grey", size = 3, fontface = "bold") +
  annotate("text", x = 152.9,  y = -28.43, label = "NEW SOUTH WALES",
           colour = "grey", size = 3, fontface = "bold") +
  geom_point(data = buoys, aes(lon, lat, fill = site),
             shape = 21, size = 4.2, stroke = 0.7, colour = "white") +
  geom_text(data = buoys, aes(lon, lat, label = site),
            hjust = 0, nudge_x = 0.04, size = 3.6, fontface = "bold") +
  scale_fill_manual(values = site_cols, guide = "none") +
  coord_sf(xlim = c(152.75, 154.05), ylim = c(-28.5, -27.55), expand = FALSE) +
  labs(x = NULL, y = NULL,
       title = "Two buoys, ~27 km apart on the same coast") +
  theme(panel.background = element_rect(fill = "lightblue", colour = NA),
        panel.grid = element_line(colour = "grey"))
```

At the **`r format(n_matched, big.mark=",")`** half-hours when both were recording, their wave heights agree to a correlation of **`r fmt2(cor_hs_sites)`** and their sea temperatures to **`r fmt2(cor_sst_sites)`**. The seasonal cycle and the storm spikes are not one float's quirks — they are *the ocean*. Tweed runs a touch bigger (mean Hs **`r fmt2(tw_mhs)`** vs **`r fmt2(gc_mhs)`** m; biggest single wave **`r fmt2(tw_hmax)`** vs **`r fmt2(gc_hmax)`** m), consistent with its more exposed position.

```{r}
#| label: fig-compare
#| fig-cap: "Daily maximum Hs at both buoys. The two lines move together all year and spike on the same days — but Gold Coast goes blank for much of March (its outage), exactly when a major swell hit. Tweed Heads, near-complete, covers the gap."
waves %>% 
  group_by(site, date) %>% 
  summarise(max_hs = max(hs, na.rm = TRUE), .groups = "drop") %>% 
  mutate(max_hs = ifelse(is.finite(max_hs), max_hs, NA_real_)) %>% 
  ggplot(aes(date, max_hs, colour = site)) +
  geom_hline(yintercept = 3, linetype = "dashed", colour = "grey70") +
  geom_line(linewidth = 0.45) +
  scale_colour_manual(values = site_cols, name = NULL) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(x = NULL, y = "Daily max Hs (m)",
       title = "Same ocean, two buoys: the lines rise and fall together") +
  theme(legend.position = "top")
```

The payoff is **March**. Gold Coast lost **`r fmt1(mar_miss)`%** of the month to its outage — right across a major event. Tweed Heads, **`r fmt1(mar_tw$pct_missing_hs)`%** missing, covered the gap and showed March peaking at **`r fmt2(mar_tw$max_hs)` m** — *rougher* than Gold Coast's broken record let on. The second buoy didn't just confirm the story; it **rescued the part the first one dropped.** That, more than any single statistic, is the argument for reaching past one dataset.

#### A trap worth pointing out

It's tempting to relate wave height to sea temperature directly, and if you do, you get a mild *positive* monthly correlation (**`r fmt2(cor_month)`**) — bigger waves in warmer months? That's a **seasonal mirage**. It only appears because autumn happens to be *both* still-warm *and* the stormiest part of 2025; there's no physical link. The two record-breaking events make the point better than any coefficient: one struck at `r fmt1(top_hmax$sst)`°C and the other at `r fmt1(top_hs$sst)`°C. Sea temperature and wave height are driven by **different machines** — the sun's slow heating versus the wind's fast storms — and reading causation into their overlap is exactly the kind of mistake a year of honest data helps you avoid.

#### What the year says

Across 2025 the buoys told a four-part story: sea temperature and wave height **march to different drummers**; the big swells **arrive in a couple of bunches** rather than spread evenly, from opposite seasons; wave **period** quietly separates local chop from ocean swell; and a second buoy 27 km south **confirms all of it at \~96% agreement** — while quietly filling the gap the first buoy left in March. None of it is a forecast — it's two buoys, one year — but it's a satisfying amount of structure to pull out of a couple of yellow floats and a year of patience.

### References

- Department of Environment, Tourism, Science and Innovation (Queensland Government). *Coastal Data System – Waves (Gold Coast): Wave data – 2025* \[data set\]. CC BY 4.0. <https://www.data.qld.gov.au/dataset/coastal-data-system-waves-gold-coast>
- Department of Environment, Tourism, Science and Innovation (Queensland Government). *Coastal Data System – Waves (Tweed Heads): Wave data – 2025* \[data set\]. CC BY 4.0. <https://www.data.qld.gov.au/dataset/coastal-data-system-waves-tweed-heads>
- Queensland Government. *Wave monitoring — field definitions & site locations.* <https://www.qld.gov.au/waves>
- Wickham, H. et al. (2019). *Welcome to the tidyverse.* JOSS 4(43), 1686. R packages `tidyverse`, `lubridate`, `scales`.
- Pebesma, E. (2018). *Simple Features for R (`sf`).* The R Journal 10(1), 439–446. Australian state boundaries via the `ozmaps` package.

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
