---
title: "When do the big waves come? A year on two neighbouring wave buoys"
subtitle: "Assignment 4 ETC5512"
author: "Hanwen Liu"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 8
    fig-height: 4.6
    fig-align: center
    fig-cap-location: bottom
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

```{r}
#| label: setup
#| include: false

source("R/02_clean_process.R", local = knitr::knit_global())

library(tidyverse)
library(scales)
library(sf)
library(ozmaps)

theme_set(
  theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.title = element_text(face = "bold"),
          plot.subtitle = element_text(colour = "grey"),
          strip.text = element_text(face = "bold"))
)
season_cols <- c(Summer = "red", Autumn = "yellow",
                 Winter = "blue", Spring = "green")
site_cols   <- c("Gold Coast" = "lightblue", "Tweed Heads" = "pink")

buoys <- tibble(
  site = factor(c("Gold Coast", "Tweed Heads"), levels = levels(waves$site)),
  lon  = c(153.44095, 153.576333),
  lat  = c(-27.9649,  -28.177083),
  depth_m = c(17, 22)
)
haversine <- function(lo1, la1, lo2, la2) {
  R <- 6371; dlat <- (la2 - la1) * pi/180; dlon <- (lo2 - lo1) * pi/180
  a <- sin(dlat/2)^2 + cos(la1*pi/180) * cos(la2*pi/180) * sin(dlon/2)^2
  2 * R * asin(sqrt(a))
}
dist_km <- haversine(buoys$lon[1], buoys$lat[1], buoys$lon[2], buoys$lat[2])

gc         <- waves   %>% 
  filter(site == "Gold Coast")
monthly_gc <- monthly %>%  
  filter(site == "Gold Coast")

warm  <- monthly_gc %>% 
  slice_max(mean_sst, n = 1)
cool  <- monthly_gc %>% 
  slice_min(mean_sst, n = 1)
rough <- monthly_gc %>% 
  slice_max(mean_hs,  n = 1)
calm  <- monthly_gc %>% 
  slice_min(mean_hs,  n = 1)
sst_swing <- warm$mean_sst - cool$mean_sst

n_big3   <- sum(gc$hs >= 3, na.rm = TRUE)
pct_big3 <- 100 * mean(gc$hs >= 3, na.rm = TRUE)
top_hs   <- gc %>% 
  slice_max(hs,   n = 1)   
top_hmax <- gc %>% 
  slice_max(hmax, n = 1)   
mar_miss <- monthly_gc$pct_missing_hs[monthly_gc$month == "Mar"]

cor_hs_tp  <- cor(gc$hs, gc$tp,  use = "complete.obs")
cor_hs_sst <- cor(gc$hs, gc$sst, use = "complete.obs")
cor_month  <- cor(monthly_gc$mean_hs, monthly_gc$mean_sst)

site_stats <- waves %>% 
  group_by(site) %>% 
  summarise(mean_hs = mean(hs, na.rm = TRUE),
            mean_sst = mean(sst, na.rm = TRUE),
            max_hmax = max(hmax, na.rm = TRUE), .groups = "drop")
gc_mhs  <- site_stats$mean_hs[site_stats$site == "Gold Coast"]
tw_mhs  <- site_stats$mean_hs[site_stats$site == "Tweed Heads"]
gc_hmax <- site_stats$max_hmax[site_stats$site == "Gold Coast"]
tw_hmax <- site_stats$max_hmax[site_stats$site == "Tweed Heads"]

wide <- waves %>% 
  mutate(s = recode(site, "Gold Coast" = "gc", "Tweed Heads" = "tw")) %>% 
  select(s, datetime, hs, sst) %>% 
  pivot_wider(names_from = s, values_from = c(hs, sst))
hsp  <- wide %>% 
  filter(!is.na(hs_gc),  !is.na(hs_tw))
sstp <- wide %>% 
  filter(!is.na(sst_gc), !is.na(sst_tw))
cor_hs_sites  <- cor(hsp$hs_gc,  hsp$hs_tw)
cor_sst_sites <- cor(sstp$sst_gc, sstp$sst_tw)
n_matched     <- nrow(hsp)
mar_tw <- monthly %>% 
  filter(site == "Tweed Heads", month == "Mar")

states <- ozmaps::ozmap_states

fmt1 <- function(x) formatC(x, format = "f", digits = 1)
fmt2 <- function(x) formatC(x, format = "f", digits = 2)
```

::: panel-tabset
<!-- ===================================================================== -->

<!-- TASK 1 -->

## Data and Documenting

### Q1 — The question

This blog post answers, from a full year of half-hourly wave-buoy records:

> **How did wave conditions and sea temperature change across 2025 off the Gold Coast, when did the biggest-wave events arrive, and how does wave height relate to wave period (swell versus local wind-sea) — and does a second buoy 27 km south tell the same story?**

I chose this question because it is tightly scoped — one year, two open datasets — yet rich enough to exercise a wide range of the unit's skills: datetime parsing, sentinel-value cleaning, aggregating a 17,520-row half-hourly series, joining two datasets on a common time grid, and time-series, seasonal, spatial and correlation analysis. Adding the second buoy is deliberate: rather than just describing one record, it lets me *test* whether the patterns are real signals in the ocean or merely one instrument's quirks.

### Q2 — The data {#sec-data}

I use **two** datasets from the **Queensland Government Open Data Portal**, both the "Wave data – 2025" resource of a *Coastal Data System – Waves* dataset, both **observational sensor data** from a **Datawell Directional Waverider** buoy recording a 26.6-minute sample every **30 minutes**.

|   | **Dataset 1 — Gold Coast** (primary) | **Dataset 2 — Tweed Heads** (comparison) |
|------------------------|------------------------|------------------------|
| Source / custodian | DETSI + City of Gold Coast | DETSI + Tweed River Sand Bypassing |
| Location | 27.965° S, 153.441° E, 17 m depth | 28.177° S, 153.576° E, 22 m depth |
| Rows (2025) | 17,520 | 17,520 (after clipping to 2025) |
| Licence | CC BY 4.0 | CC BY 4.0 |
| Portal | [link](https://www.data.qld.gov.au/dataset/coastal-data-system-waves-gold-coast) | [link](https://www.data.qld.gov.au/dataset/coastal-data-system-waves-tweed-heads) |

**Type.** Neither is a survey or an experiment; each is effectively a **census** of every valid half-hourly record the buoy made in 2025 — a complete, regular time grid.

**Why this suits the question.** Each file carries exactly the variables I need — significant and maximum wave height (`Hs`, `Hmax`), two wave periods (`Tz`, `Tp`), peak wave direction, and sea surface temperature (`SST`). One buoy answers the "when" questions; the **second, 27 km south, is what lets me check whether the answers are real**. Both are \~1 MB and openly licensed, so the work is fully reproducible.

**Limitations.**

- **Missing data is coded, not blank.** Failed readings are the sentinel `-99.9`; untreated it would poison every average. Gold Coast is \~4.5% missing for wave parameters — but **not evenly**: **March alone is \~`r fmt1(mar_miss)`% missing** (an outage). Tweed Heads is far more complete (\~0.1%), which turns out to matter.
- **Two buoys, one year.** Findings describe these sites in 2025, not a long-term climatology.
- **Point in space.** A buoy measures open-water sea state, not the breaking surf at any particular beach.

**Ethics & privacy.** Both datasets measure the physical ocean only — **no people, vessels or identifiable subjects** — so privacy/re-identification risk is negligible. The obligations are *attribution* (CC BY 4.0) and *responsible interpretation*: this is a curiosity project, **not** a marine-safety product.

### Q3 — Downloading and processing the data

**Download (`R/01_download.R`).** The script downloads each buoy's CSV from its portal URL and **falls back to the committed copy in `data/raw/`** if the portal is unreachable, so the pipeline reproduces online *or* offline. Raw files are never edited by hand.

```{r}
#| eval: false
#| code-fold: false

ok <- tryCatch({ download.file(s$url, s$dest, mode = "wb"); TRUE },
               error = function(e) FALSE, warning = function(w) FALSE)
if (!ok && !file.exists(s$dest)) stop("Download failed and no committed copy.")
header <- strsplit(readLines(s$dest, n = 1), ",")[[1]]
stopifnot(identical(header, expected_cols))
```

**Process (`R/02_clean_process.R`).** The core steps — applied to both buoys:

```{r}
#| eval: false
#| code-fold: false

read_buoy <- function(path, site) {
  read_csv(path, col_types = cols(`Date/Time (AEST)` = col_character(),
                                  .default = col_double())) %>% 
    mutate(site = site, .before = 1)
}
waves <- bind_rows(read_buoy(gc_path, "Gold Coast"),
                   read_buoy(tw_path, "Tweed Heads")) %>% 
  mutate(datetime = ymd_hms(`Date/Time (AEST)`, truncated = 1, tz = "Australia/Brisbane")) %>% 
  mutate(across(c(hs, hmax, tz, tp, dir, sst), \(x) if_else(x == -99.9, NA, x))) %>% 
  filter(year(datetime) == 2025)  
```

The pipeline then runs **per-site assertions** (each site: 17,520 rows, every timestamp parsed, no duplicates, a regular 30-minute grid) and writes two **processed** datasets: `waves_2025_clean.csv` (tidy half-hourly, both sites, sentinel → `NA`) and `waves_2025_monthly.csv` (site × month summary). Raw and processed data are both saved under `data/`.

### Q4 — Data documentation for markers

Full documentation lives alongside the data:

- **`docs/metadata.md`** — provenance, custodians, licences, coverage, instruments, ethics (both buoys).
- **`docs/data_dictionary.csv`** — every variable with units, valid 2025 range and missing code.
- **`README.md`** — folder map and one-command reproduction steps.

A **sample of a raw file** (note the `-99.9` sentinel and the half-hourly step):

```{r}
#| code-fold: true
#| code-summary: "Show code"
read_csv("data/raw/gold-coast-mk4-2025-waveverifieddata.csv", n_max = 3,
         col_types = cols(.default = col_character())) |>
  knitr::kable(caption = "First 3 rows of the Gold Coast raw CSV.")
```

**Missingness by month and site** — the one big gap is Gold Coast in March; Tweed Heads is near-complete all year (this is what makes it a useful cross-check):

```{r}
#| label: tbl-missing
#| code-fold: true
#| fig-height: 3.0
#| fig-cap: "Percentage of half-hours with no valid Hs reading, by month and buoy. Gold Coast's ~25% March outage is the one large gap; Tweed Heads is near-complete all year."
monthly |>
  ggplot(aes(month, pct_missing_hs, fill = site)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = site_cols, name = NULL) +
  labs(x = NULL, y = "% missing Hs",
       title = "Where are the gaps?  Missing wave readings by month, 2025") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  theme(legend.position = "top")
```

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
